Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study

### 2024-03-19 manuscript.Rmd
19/3

This was a registry-based study using all patients between 2014 and 2022
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bi-variable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, Revised Trauma Score
(RTS), Injury Severity Score (ISS), on call, time from arrival at the
hospital until first computed tomography (CT), days in the ICU, ASA
preinjury and survival after 30 days.

@@

OFI was identified in 143 of 1449 patients. Revised Trauma Score, ≤ 7
days in the ICU and death within 30 days after hospitalization were
significantly associated with OFI in the unadjusted analysis, and
Revised Trauma Score and ≤ 7 days in the ICU were significantly
associated with OFI in the adjusted analysis.

@@

For the RTS-value the GCS, SBP and RR values were converted to a
RTS-code between 0 and 4.

For patients where hospital values for systolic blood pressure,
respiratory rate or GCS score were missing, prehospital values were used
instead. If the patient was intubated A complete case analysis was
conducted after handling missing values. We present sample
characteristics using descriptive statistics.

@@

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS as mentioned in the
introduction. An explanation for the increased odds of OFI in patients
with increased RTS could be that the patients with low RTS values - with
low GCS, RR and SBP also are the once that are in a bad state upon
arrival - upon which emergency procedures e.g. intubation have been
carried out rapidly due to the patients being very ill from the
beginning. The patients who are being missed and where certain
procedures take longer time before being performed are the patients who
seem stable upon arrival, but who later become worse --- the patients
with higher RTS scores. In Albaaj's study respiratory rate \>29,
systolic blood pressure 50-75 mmHg and 76-89 mmHg, and Glasgow Coma
Scale 9-12 were significantly related to increased odds of OFI in the
unadjusted analysis, but not in the adjusted analysis [@12]. All of
these variables equal medium RTS values.

The variable ≤ 7 days in the ICU was significantly associated with
decreasing odds of OFI in both the adjusted and unadjusted analyses. An
explanation would be that less days spent in the ICU compared to more
days spent in the ICU indicates that patients have been in a better
state and needed less time of recovery before leaving the hospital.
Furthermore, a longer stay leads to more space for mistakes being made.

@@

@@

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna. This leads to a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI later on in the hospitalization
process. #Lägga in källa för audit filters - Kelvins artikel?

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes
because we included only those who were admitted to the ICU, \> 14 years
old, not dead on arrival and had a registration for the presence or
absence of OFI. The external validity is therefore compromised due to
the limited size of the Swedish database.

@@

The aim of this study was to characterize patient and process factors
among adult patients admitted to the ICU that were associated with OFI.
The study showed that higher RTS score is associated with increased odds
of OFI indicating that patients who are relatively well upon arrival to
the hospital are the group where improved trauma care should be focused
on. The logistical regression also showed that ≤ 7 days in the ICU was
associated with decreased odds of OFI indicating that patients who are
at the hospital less days in the ICU compared to more days in the ICU
are less prone to opportunities for improvement.

>>>>

This was a registry-based study using all patients between 2014 and 2023
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bi-variable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, Revised Trauma Score
(RTS), Injury Severity Score (ISS), on call, time from arrival at the
hospital until first computed tomography (CT), days in the ICU, ASA
preinjury and survival after 30 days.

@@

OFI was identified in 143 of 1449 patients. Higher Revised Trauma Score
(RTS) was associated with higher odds of OFI, ≤ 7 days in the intensive
care unit (ICU) was associated with lower odds of OFI, and death within
30 days after hospitalization was associated with higher odds of OFI
with statistical significance in the unadjusted analysis. Higher Revised
Trauma Score (RTS) was associated with higher odds of OFI, and ≤ 7 days
in the ICU was associated with lower odds of OFI with statistical
significance in the adjusted analysis.

@@

For patients where hospital values for systolic blood pressure,
respiratory rate or GCS score were missing, prehospital values were used
instead. If the patient was intubated we automatically assumed that GCS
was equal to 3 and RR was equal to 0. For the RTS-variable, each of the
variables GCS, SBP and RR were then converted to a RTS-code between 0
and 4 based on what the values of GCS, SBP and RR were respectively. RTS
was then calculated using the formula for RTS based on the combination
of the RTS-codes assigned to each of the previously listed variables.

A complete case analysis was conducted after handling missing values. We
present sample characteristics using descriptive statistics.

@@

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS. An explanation for the
increased odds of OFI in patients with increased RTS could be that the
patients with low RTS values - with low GCS, RR and SBP also are the
once that are in a bad state upon arrival --- upon which emergency
procedures e.g. intubation have been carried out rapidly due to the
patients being visibly ill from the beginning. The patients who are
being missed and where certain procedures take longer time before being
performed are the patients who seem stable upon arrival, but who later
become worse --- the patients with higher RTS scores. In Albaaj's study
respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
mmHg, and Glasgow Coma Scale 9-12 were significantly related to
increased odds of OFI in the unadjusted analysis, but not in the
adjusted analysis among patients from the Swedish Trauma Registry [@12].
All of these variables equal medium to higher RTS values --- capturing
the group of patients who's vital parameters are negatively affected by
the trauma, but who are relatively stable in this group of patients.
This creates a sense of false security, leading to increased time until
necessary emergency procedures are performed.

#Se om det finns källa som pekar på vilken grupp av patienter som bör få
mer fokus på IVA?

The variable ≤ 7 days in the ICU was significantly associated with
decreasing odds of OFI in both the adjusted and unadjusted analyses.
There are two aspects to consider concerning this variable. One aspect
would be that less days spent in the ICU compared to more days spent in
the ICU indicates that patients have been in a better state and needed
less time of recovery before leaving the hospital. Furthermore, a longer
stay leads to more space for mistakes being made. The other aspect of
the variable would be that few days in the ICU could indicate the
patient passing away before any mistakes having been made due to the
patient being very ill from the beginning --- the deaths were not
preventable, but rather expected.

#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
vad orsakerna är

@@

Sex, age and ASA preinjury were not significant in the unadjusted nor
the adjusted analyses, which implicates that the patients were not
treated differently based on age or gender. The results concerning the
ASA preinjury could be due to the OFI considering if each patient really
had an opportunity for improvement based on their individual cases,
meaning that if a patient has a higher ASA-score before the trauma, that
patient may not have the same reference point for OFI compared to a
patient with low ASA-score before the trauma.

#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
bedömningen av OFI

The variable on duty was not significant in either analyses indicating
that the time of arrival of the patients did not affect their medical
treatment. Trying to reduce number of OFI can therefore not be solved by
increasing medical resources outside working hours.

#Lägga in artikel som pekar på om duty påverkar handläggning eller inte

Something to consider is why ISS was not significant in the analyses.
Patients admitted to the ICU are usually in a bad shape with generally
high ISS. ISS is based on the square of the three highest AIS scores
allocated to the three most severly injured body parts. The downside of
ISS on its own is that it has higher predictive value when combined with
other parameters like vital signs, comorbidities and mechanism of
injury.

#Artikel om hur ISS bedöms kunna användas i samanhang av trauma

@@

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna. This leads to a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI later on in the hospitalization
process.

#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes
because we included only those who were admitted to the ICU, \> 14 years
old, not dead on arrival and had a registration for the presence or
absence of OFI. The external validity is therefore compromised due to
the limited size of the Swedish trauma database of patients admitted to
the ICU.

@@

The aim of this study was to characterize patient and process factors
among adult patients admitted to the ICU that were associated with OFI.
The study showed that higher RTS score is associated with increased odds
of OFI indicating that patients who are relatively well upon arrival to
the hospital are the group where improved trauma care should get more
focus than they do now. The logistical regression also showed that ≤ 7
days in the ICU was associated with decreased odds of OFI indicating
that patients who are at the hospital less days in the ICU compared to
more days in the ICU are less prone to opportunities for improvement.
